Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-10 appendix.tex
first rough draft

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for
$\beta\gtrsim1$, and as $\beta\to\infty$ the VGSB framework gives a
(spurious) positive growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
correct behavior  ($\nu\to0$ as $\beta\to\infty$).

@@

Fig. \ref{gcorr_compare} shows that provided we consider $\beta\ll1$,
then the VGSB framework is an adequate approximation. It is
interesting to note that there is a thermal relaxation timescale that
maximizes the mode decay rate. Here, it is $\beta\simeq2$ or about
$0.3$ orbital periods. This is, in fact, consistent with Fig. 12 of
\cite{nelson13}.

@@

\section{Eigenfrequency in locally
  isothermal disks in global cylindrical geometry}
Consider a locally isothermal disk in 3D such that $P=c_s^2\rho$ at
all times, so in the perturbed state $\delta P =
c_s^2\delta\rho$, with $c_s(r)$ begin a prescribed function. 
The linearized fluid equations in cylindrical
geometry is then 
\begin{align}
 \ii\sigma \delta v_r + 2\Omega\delta v_\phi &= c_s^2\frac{\p}{\p r}\left(\frac{\delta\rho}{\rho}\right),\\
 \ii\sigma \delta v_\phi &= \frac{1}{r}\delta\bm{v}\cdot\nabla j,\\
 \ii\sigma \delta v_z &= c_s^2\frac{\p}{\p
   z}\left(\frac{\delta\rho}{\rho}\right),\\
 \ii\sigma\delta\rho &= \nabla\cdot\left(\rho\delta\bm{v}\right).
\end{align}
We can eliminate $\delta v_\phi$ from these equations and obtain
\begin{align}
  -\sigma^2\rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) +
  \kappa^2\rho|\delta v_r|^2 + \frac{2\Omega}{r}\frac{\p j}{\p z} \rho
  \delta v_r^*\delta v_z =
  c_s^2\rho\delta\bm{v}^*\cdot\nabla\left(\frac{\ii\sigma\delta\rho}{\rho}\right)
  =c_s^2\rho\delta\bm{v}^*\cdot\nabla\left[\frac{1}{\rho}\nabla\cdot\left(\rho\delta\bm{v}\right)\right].\label{global_integral}
\end{align}
At this point if we make the anelastic approximation by assuming
$\nabla\cdot\left(\rho\delta\bm{v}\right)$ is negligible in this equation \citep{nelson13},
then the only contribution to instability comes from the third term on
the left hand side, i.e. the explicit vertical shear term $\p_zj$. The 
term proportional to $\kappa^2$ is stabilizing. Because radial shear
is much larger than vertical shear, this implies that significant
instability can only be achieved with large vertical
motions (i.e. $|\delta v_z|\gg|\delta v_r|$).

If we integrate Eq. \ref{global_integral} over a volume
and assume boundary terms vanish when integrating by parts, we obtain 
\begin{align}
  -\sigma^2\int \rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) dV +
  \int \kappa^2\rho|\delta v_r|^2 dV + \int \frac{2\Omega}{r}\frac{\p j}{\p z} \rho
  \delta v_r^*\delta v_z dV = -\int
  \frac{c_s^2}{\rho}\left|\nabla\cdot\left(\rho\delta\bm{v}\right)\right|^2
  dV - \int \nabla\cdot\left(\rho\delta\bm{v}\right) \delta
  \bm{v}^*\cdot\nabla c_s^2.
\end{align}
Recalling $j=r^2\Omega$ and $r\p_z\Omega^2 = -\p_z\ln{\rho}\p_r c_s^2$
for a locally isothermal disk, we obtain
\begin{align}
  \sigma^2\int \rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) dV =
  \int \kappa^2\rho|\delta v_r|^2 dV + \int
  \frac{c_s^2}{\rho}\left|\nabla\cdot\left(\rho\delta\bm{v}\right)\right|^2
  dV + \int \left[ \rho \left(\nabla\cdot\delta\bm{v}\right)\delta
    v_r^* + |\delta v_r|^2\frac{\p \rho}{\p r}\right]\frac{dc_s^2}{dr}  dV.
\end{align}
We see that instability is only possible if
$dc_s^2/dr\neq0$. Furthermore, since density and temperature typically
decrease outwards in astrophysical disks, instability is associated
with the term proportional to $\nabla\cdot\delta\bm{v}$. It is also
clear that $\sigma^2$ will be complex in general.

>>>>

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the VGSB framework gives a
(spurious) positive growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
correct behavior  ($\nu\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider
$\beta\ll1$, then the VGSB framework is an adequate approximation.

@@

@@

\section{Axisymmetric instability in vertically 
  isothermal disks in global cylindrical geometry}\label{global_vertiso}
Consider a locally isothermal disk in 3D such that $P=c_s^2\rho$ at
all times, so in the perturbed state $\delta P =
c_s^2\delta\rho$, with $c_s$ begin a prescribed function of position. 
The linearized fluid equations in cylindrical
geometry is then 
\begin{align}
 \ii\sigma \delta v_r + 2\Omega\delta v_\phi &= c_s^2\frac{\p}{\p r}\left(\frac{\delta\rho}{\rho}\right),\\
 \ii\sigma \delta v_\phi &= \frac{1}{r}\delta\bm{v}\cdot\nabla j,\\
 \ii\sigma \delta v_z &= c_s^2\frac{\p}{\p
   z}\left(\frac{\delta\rho}{\rho}\right),\\
 \ii\sigma\delta\rho &= \nabla\cdot\left(\rho\delta\bm{v}\right).
\end{align}
We can eliminate $\delta v_\phi$ from these equations to obtain
\begin{align}
  -\sigma^2\rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) +
  \kappa^2\rho|\delta v_r|^2 + \frac{2\Omega}{r}\frac{\p j}{\p z} \rho
  \delta v_r^*\delta v_z =
  c_s^2\rho\delta\bm{v}^*\cdot\nabla\left(\frac{\ii\sigma\delta\rho}{\rho}\right)
  =c_s^2\rho\delta\bm{v}^*\cdot\nabla\left[\frac{1}{\rho}\nabla\cdot\left(\rho\delta\bm{v}\right)\right].\label{global_integral} 
\end{align}
At this point if we make the anelastic approximation by assuming
$\nabla\cdot\left(\rho\delta\bm{v}\right)$ is negligible in this
equation \citep[as argued by][]{nelson13}, 
then the only contribution to instability comes from the third term on
the left hand side, i.e. the explicit vertical shear term
$\p_zj$. This is the same conclusion reached in
\S\ref{integral_relation}.

The term proportional to $\kappa^2$ is stabilizing. Because radial shear
is much larger than vertical shear, this implies that significant
instability can only be achieved with large vertical
motions (i.e. $|\delta v_z|\gg|\delta v_r|$).

If we integrate Eq. \ref{global_integral} over the fluid volume 
and assume boundary terms vanish when integrating by parts, we obtain 
\begin{align}
  -\sigma^2\int \rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) dV +
  \int \kappa^2\rho|\delta v_r|^2 dV + \int \frac{2\Omega}{r}\frac{\p j}{\p z} \rho
  \delta v_r^*\delta v_z dV = -\int
  \frac{c_s^2}{\rho}\left|\nabla\cdot\left(\rho\delta\bm{v}\right)\right|^2
  dV - \int \nabla\cdot\left(\rho\delta\bm{v}\right) \delta
  \bm{v}^*\cdot\nabla c_s^2.
\end{align}
Note that the treatment so far is in fact valid for any prescribed
$c_s(r,z)$. However, to proceed further we specialize to the case
$c_s=c_s(r)$ for simplicity, i.e. vertically isothermal
disks. Recalling $j=r^2\Omega$ and $r\p_z\Omega^2 = -\p_z\ln{\rho}\p_r
c_s^2$  for a vertically isothermal disk, we obtain 
\begin{align}
  \sigma^2\int \rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) dV =
  \int \kappa^2\rho|\delta v_r|^2 dV + \int
  \frac{c_s^2(r)}{\rho}\left|\nabla\cdot\left(\rho\delta\bm{v}\right)\right|^2
  dV + \int \left[ \rho \left(\nabla\cdot\delta\bm{v}\right)\delta
    v_r^* + |\delta v_r|^2\frac{\p \rho}{\p r}\right]\frac{dc_s^2(r)}{dr}  dV.
\end{align}
We see that instability is only possible if
$dc_s^2/dr\neq0$. Furthermore, since density and temperature typically
decrease outwards in astrophysical disks, instability is associated
with the term proportional to $\nabla\cdot\delta\bm{v}$. It is also
clear that $\sigma^2$ will be complex in general.

### 2015-02-10 numerical.tex
paper updates, code updates
first rough draft

\section{Numerical solutions}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the thin-disk, low-frequency and nearly-Keplerian
approximations used in the analytical discussion above, and  
use the full expressions of $\rho$, $\Omega$ and
$\kappa$ given in \S\ref{eqm}. We also restore $\hat{g}_c=1$ to
account for the background radial disk structure.

@@

Fig. \ref{bcrit_compare1} shows the growth rate $\nu$ of the
fundamental VSI as a function of $\beta$ for $\khat\in[1,10^2]$. We
also plot the critical thermal relaxation timescale
$\beta_\mathrm{crit}$ given by Eq. \ref{iso_vsi_cond}. For the present
disk parameters $\beta_\mathrm{crit}=0.125$.

@@

@@

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/bcrit_compare} 
   \caption{Growth rate of the fundamental VSI as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The vertical line
     is the critical thermal timescale $\beta_\mathrm{crit}$ for the
     fundamental mode according to Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/gcorr_compare_vnoniso}
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation timescale $\beta$, in vertically
    non-isothermal disks with $\Gamma=1.4$ and
    $\gamma\in[1.4,2.5]$. The disk is neutrally
    stratified for $\gamma=1.4$ and stably stratified for
    $\gamma>1.4$. Other disk parameters are
    $(p,q,\epsilon)=(0,-1,0.05)$ and the perturbation wavenumber is
    $\khat=30$.  
    \label{gcorr_compare_vnoniso}}
\end{figure}

>>>>

\section{Numerical solutions}\label{numerical}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the thin-disk, low-frequency and nearly-Keplerian
approximations used in the analytical discussion above, and  
use the full expressions of $\rho$, $\Omega$ and
$\kappa$ given in \S\ref{eqm}. We also restore $\hat{g}_c=1$ to
account for the background radial disk structure.

@@

Fig. \ref{bcrit_compare1} shows the growth rate $\nu$ of the
fundamental VSI as a function of $\beta$ for $\khat\in[1,10^2]$. We
also plot the critical thermal relaxation timescale
$\beta_\mathrm{crit}=0.125$ given by Eq. \ref{iso_vsi_cond} for the present
disk parameters.

@@

It is interesting to note that there is a thermal relaxation timescale that
maximizes the mode decay rate for $1\lesssim \khat \lesssim 30$. Here,
it is $\beta\simeq1$---$2$ or about $0.1$---$0.3$ orbital
periods. This is in fact consistent with Fig. 12 of \cite{nelson13}
where the perturbed kinetic energy in their numerical simulations was
observed to decay most rapidly for a thermal timescale of $\sim 0.1$
orbits.

@@

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{Growth rate of the fundamental VSI as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The vertical line
     is the critical thermal timescale $\beta_\mathrm{crit}$ for the
     fundamental mode according to Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/gcorr_compare_vnoniso2}
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation timescale $\beta$, in vertically
    non-isothermal disks with $\Gamma=1.4$ and
    $\gamma\in[1.4,2.5]$. The disk is neutrally
    stratified for $\gamma=1.4$ and stably stratified for
    $\gamma>1.4$. Other disk parameters are
    $(p,q,\epsilon)=(0,-1,0.05)$ and the perturbation wavenumber is
    $\khat=30$.  
    \label{gcorr_compare_vnoniso}}
\end{figure}

### 2015-02-10 paper.tex
first rough draft

%recent work on vsi 
One possibiliy is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. The VSI is the
disk analog of the Goldreich-Schubert-Fricke instability
\citep[GSF,]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple explanation for this 
instability: an exchange of fluid parcels while conserving their
angular momenta individually, may result in a new configuration of
lower energy, if the exchange occurs over a sufficiently large
vertical displacement.

@@

The VSI has been studied analytically in local approximation
\citep{urpin98,urpin03}, where a Fourier analysis was performed in 
both the radial and vertical directions. Simulations of the  
VSI have been performed by \cite{arlt04}, and more recently by 
\cite{nelson13}. In these simulations the authors indeed observe the
growth of vertically-global but radially-localized perturbations when
the equilibrium rotation varied with height.

@@

From these works it is clear that the VSI is sensitive 
to the thermodynamic response of the disk except in the special case of neutrally
stratified disks. It is therefore important to understand this dependence in order to evaluate
the viability of the VSI in realistic disks. However,

%what we're doing

@@

>>>>

%recent work on vsi 
One possibiliy is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. The VSI is the
disk analog of the Goldreich-Schubert-Fricke instability
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple explanation for this 
instability: an exchange of fluid parcels while conserving their
angular momenta individually, may result in a new configuration of
lower energy, if the exchange occurs over a sufficiently large
vertical displacement.

@@

The VSI has been studied analytically in local approximation
\citep{urpin98,urpin03}. Simulations of the  
VSI have been performed by \cite{arlt04}, and more recently by 
\cite{nelson13}. In these simulations the authors indeed observe the
growth of vertically-global but radially-localized perturbations when
the equilibrium rotation varied with height.

@@

The VSI is sensitive to disk thermodynamics except
in the special case of neutrally stratified disks. It is therefore
important to understand and quantify this dependence, for example on
disk parameters, in order to evaluate the viability of the VSI in
realistic disks. Numerical simulations of the VSI are challenging,
however, because global disk simulations should be carried out while
resolving radial scales much smaller than the local disk scale height.

On the other hand, should a given disk be
unstable to the VSI, it must pass through the linear instability
first. Thus, studying the thermodynamic dependence of the linear VSI
is a useful first step. Furthermore, linear stability calculations are
essential to complement future numerical simulations.

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations between varied from isothermal to adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of early
analysis of \cite{urpin98} and \cite{urpin03} who consider
vertically-localized disturbances. Our vertically-global approach is
necessary to study disturbances with vertical wavelengths
comparable to or larger than the disk scaleheight. We thus focus on
vertically-global perturbations, which were found by \cite{nelson13}
to dominate their numerical simulations.

@@

%say "grossly simplified

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially-local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic thermal timescale below which
the VSI is expected to operate. In \S\ref{numerical} we numerically
solve the linearized fluid equations in the radially-local
approximation. We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebulae (MMSN) and find that the VSI may operate in the outer
parts of the MMSN. We summarize and discuss our results in
\S\ref{summary}, including caveats and possible directions for future
work.

### 2015-02-10 setup.tex
first rough draft

\section{Governing equations and disk models}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$. We adopt cylindrical
co-ordinates $(r,\phi, z)$ centered on the star to describe the global
disk. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    \left.\frac{P}{\rho}\right|_{t=0}\right), \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq. \ref{full_energy}) we have
included a thermal relaxation term which restores the disk temperature
$T\propto P/\rho$ to its initial value on a characteristic timescale
$t_c=\beta\Omega_k^{-1}$ where $\beta$ is a dimensionless coefficient
and $\Omega_k=\sqrt{GM_*/r^3}$ is the Keplerian frequency.

@@

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p R} \right)&>0.  
\end{align}
\citep{tassoul78}, where $S = c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy, $c_p$ is the heat capacity at constant
pressure (assumed to be constant), and $j\equiv r^2\Omega$ is the
specific angular momentum.

For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we evaluate the second criterion
for the above disk models in $z>0$ where $\p_zP<0$, without loss of
generality,   
\begin{align}\label{solberg2}
  \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
    -\left(\gamma-\Gamma\right)\frac{\p\ln{P}}{\p\ln{r}} \right]. 
\end{align} 
For typical model parameters the right hand side is
$O(\epsilon^2)$. Hence, to have instability in the adiabatic limit 
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations.

\section{Linear problem} 
We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(z) + \delta\rho(r,z)\exp{\left( - \ii\sigma
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\sigma = \omega + \ii
\nu$ is a (generally) complex frequency, with $\omega$ amd $\nu$ being
the real frequency and growth rate, respectively.

@@

\subsection{Radially-local approximation}  
We assume perturbations have spatial dependence in the form
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r} 
\end{align}
such that $|k_rr|\gg1$, so that curvature terms can be ignored. For
these small radial lengthscale perturbations we also ignore the radial
variations in the background structure and radial gradients, and
evaluate all cofficients in the above equations at the fiducial radius
$r=r_0$. However, the full vertical dependence of the coefficients are
retained.

In order to emphasize on the radially-local approximation, we
introduce the local Cartesian co-ordinate $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x.\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$.   
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study of the vertical shear instability, but for analytical discussion
we will make further simplifications.

>>>>

\section{Governing equations and disk models}\label{setup}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$. We adopt cylindrical
co-ordinates $(r,\phi, z)$ centered on the star to describe the global
disk. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    \left.\frac{P}{\rho}\right|_{t=0}\right), \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq. \ref{full_energy}) we have
included a thermal relaxation term which restores the disk temperature
$T\propto P/\rho$ to its initial value on a characteristic timescale
$t_c=\beta\Omega_k^{-1}$ where $\beta$ is a dimensionless coefficient
and $\Omega_k=\sqrt{GM_*/r^3}$ is the Keplerian frequency.

@@

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $S = c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy, $c_p$ is the heat capacity at constant
pressure (assumed to be constant), and $j\equiv r^2\Omega$ is the
specific angular momentum.

For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we consider the second
criterion. For our disk models,
\begin{align}
  &\frac{\p S}{\p r} = c_p\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right],\\
  &\frac{\p S}{\p z} = c_p
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z}. 
\end{align} 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\Omega_k^2$ but
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
\end{align} 
For typical model parameters the right hand side is
$O(\epsilon^2)$. Hence, instability in the adiabatic limit 
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations.

\section{Linear problem}\label{linear}
We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(z) + \delta\rho(r,z)\exp{\left( - \ii\sigma
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\sigma = \omega + \ii
\nu$ is a (generally) complex frequency, with $\omega$ amd $\nu$ being
the real frequency and growth rate, respectively.

@@

\subsection{Radially-local approximation}  
We consider perturbations with small radial lengthscales 
so that we can ignore the radial variations in the background
structure and radial gradients. We thus evaluate all cofficients in
the above equations at the fiducial radius $r=r_0$. However, the full
vertical dependence of the coefficients are retained. Then we can
assume perturbations have spatial dependence in the form
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r},
\end{align}
and we assume $|k_rr|\gg1$ so that curvature terms can also be
ignored.

In order to emphasize the radially-local approximation, we
introduce the local Cartesian co-ordinate $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x.\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$.   
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications.
